Reasoning Graphically about Quantum Computation

### 2008-02-25 papers/reasoning-graphically/paper.tex
couple more diags
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1362 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
minor changes
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1366 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

To model the behaviour of quantum systems certain additional equations
must be satisfied.  Each family $(\delta, \epsilon)$ must form a
\emph{classical structure}
\cite{Coecke2006Quantum-Measure,Coecke2006POVMs-and-Naima}:  the pair
$(\delta, \epsilon)$  should form a cocommuative comonoid, the pair
$(\delta^\dag,\epsilon^\dag)$ must form a commutative monoid, and
together they must satisfy the isometry and frobenius equations. In
addition the family $\alpha$ must form an abelian group with $\alpha^\dag
= -\alpha$.  These equations are presented graphically below for the
green family;  the red vertices obey all the same equations
\begin{description}
\item[Comonoid Laws] 
\[
\begin{array}{ccccccccccccccc}
  \inlinegraphic{3.5em}{comonoid-assoc1}  &=\;&
  \inlinegraphic{3.5em}{comonoid-assoc2}
  &\qquad\qquad&
  \inlinegraphic{3.7em}{comonoid-unit1}  &=\;&
  \inlinegraphic{3.7em}{comonoid-unit2}  &=\;&
  \inlinegraphic{3.7em}{comonoid-unit3} 
  &\qquad\qquad&
  \inlinegraphic{3.5em}{comonoid-comm1}  &=\;&
  \inlinegraphic{3.5em}{comonoid-comm2}
\end{array}
\]
\item[Monoid Laws] 
\[
\begin{array}{ccccccccccccccc}
  \inlinegraphic{3.5em}{monoid-assoc1}  &=\;&
  \inlinegraphic{3.5em}{monoid-assoc2}
  &\qquad\qquad&
  \inlinegraphic{3.7em}{monoid-unit1}  &=\;&
  \inlinegraphic{3.7em}{monoid-unit2}  &=\;&
  \inlinegraphic{3.7em}{monoid-unit3} 
  &\qquad\qquad&
  \inlinegraphic{3.5em}{monoid-comm1}  &=\;&
  \inlinegraphic{3.5em}{monoid-comm2}
\end{array}
\]
\item[Isometry] 
\item[Frobenius]
  \[
  \begin{array}{ccccc}
    \inlinegraphic{3.5em}{frobenius1}  &=\;&
    \inlinegraphic{3.5em}{frobenius2} &=\;&
    \inlinegraphic{3.5em}{frobenius3}
  \end{array}
  \]
  \item[Abelian Unitary Group]
\end{description}

@@

In graph patterns, variable nodes generalise the concept of a
half-edge. Variable nodes, unlike boundary-nodes in the concrete
graph, are not classified into domain and co-domain. The intuition of
a variable node is that it can be replaced by any concrete node. Thus,
unlike nodes in conrete a graph, variable node's do not define an
ordering on the edges that are incident. Instead, the edges that touch
a variable node are simply a set.

The type of a variable nodes is thus defined as a multiset rather than
a tensor. Any tensor type can be lifted into a multiset type in the
obvious manner, which we denote by the operation {\tt
  mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor b
  \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type, $T$,
is an instance of a multiset type, $M$, when it's lifting into a
multiset is a super-set of $M$: $ M \subseteq
\mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
here is that an instance of variable node is allowed to have as many
extra edges as it likes as long as it at least has the ones mentioned
in the original graph. A variable node, $a$ is be said to an instance
of another variable node $b$ when the multiset represnting the type of
$a$ is a superset of the multiset for the type of $b$. Thus, with
regard to the underlying semantics, $a$ being a subtype of $b$ means
that the set of instances of $a$ are a subset of the instances of $b$.

@@

In this section, we define a compositonality principle for graphs
patterns, that is graphs with variable-nodes and !-boxes. This is
related to compositionality of the underlying concrete graphs by the
following instantiation homomorphism:

>>>>

To model the behaviour of quantum systems certain additional equations
must be satisfied.  Each family $(\delta, \epsilon)$ must form a
\emph{classical structure}
\cite{Coecke2006Quantum-Measure,Coecke2006POVMs-and-Naima}:  the pair
$(\delta, \epsilon)$  should form a cocommuative comonoid, the pair
$(\delta^\dag,\epsilon^\dag)$ must form a commutative monoid, and
together they must satisfy the isometry and frobenius equations. In
addition the family $\alpha$ must form an abelian group with $\alpha^\dag
= -\alpha$.  These equations are presented graphically below for the
green family;  the red vertices obey all the same equations
\begin{description}
\item[Comonoid Laws] 
\[
\begin{array}{ccccccccccccccc}
  \inlinegraphic{3em}{comonoid-assoc1}  &=\;&
  \inlinegraphic{3em}{comonoid-assoc2}
  &\qquad\qquad&
  \inlinegraphic{3em}{comonoid-unit1}  &=\;&
  \inlinegraphic{3em}{comonoid-unit2}  &=\;&
  \inlinegraphic{3em}{comonoid-unit3} 
  &\qquad\qquad&
  \inlinegraphic{3em}{comonoid-comm1}  &=\;&
  \inlinegraphic{3em}{comonoid-comm2}
\end{array}
\]
\item[Monoid Laws] 
\[
\begin{array}{ccccccccccccccc}
  \inlinegraphic{3em}{monoid-assoc1}  &=\;&
  \inlinegraphic{3em}{monoid-assoc2}
  &\qquad\qquad&
  \inlinegraphic{3em}{monoid-unit1}  &=\;&
  \inlinegraphic{3em}{monoid-unit2}  &=\;&
  \inlinegraphic{3em}{monoid-unit3} 
  &\qquad\qquad&
  \inlinegraphic{3em}{monoid-comm1}  &=\;&
  \inlinegraphic{3em}{monoid-comm2}
\end{array}
\]
\item[Isometry]
\[
\begin{array}{ccc}
  \inlinegraphic{3.5em}{isometry1}  &=\;&
  \inlinegraphic{3.5em}{isometry2}  
\end{array}
\]
\item[Frobenius]
  \[
  \begin{array}{ccccc}
    \inlinegraphic{3em}{frobenius1}  &=\;&
    \inlinegraphic{3em}{frobenius2} &=\;&
    \inlinegraphic{3em}{frobenius3}
  \end{array}
  \]
  \item[Abelian Unitary Group]
\end{description}

@@

In graph patterns, variable nodes generalise the concept of a
half-edge. Variable nodes, unlike boundary-nodes in the concrete
graph, are not classified into domain and co-domain. The intuition of
a variable node is that it can be replaced by any concrete node. Thus,
unlike nodes in conrete a graph, variable nodes do not define an
ordering on the edges that are incident. Instead, the edges that touch
a variable node are simply a set.

The type of a variable nodes is thus defined as a multiset rather than
a tensor. Any tensor type can be lifted into a multiset type in the
obvious manner, which we denote by the operation {\tt
  mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor b
  \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type, $T$,
is an instance of a multiset type, $M$, when it's lifting into a
multiset is a super-set of $M$: $ M \subseteq
\mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
here is that an instance of a variable node is allowed to have as many
extra edges as it likes as long as it at least has the ones mentioned
in the original graph. A variable node, $a$ is be said to an instance
of another variable node $b$ when the multiset representing the type of
$a$ is a superset of the multiset for the type of $b$. Thus, with
regard to the underlying semantics, $a$ being a subtype of $b$ means
that the set of instances of $a$ are a subset of the instances of $b$.

@@

In this section, we define a compositonality principle for graphs
patterns, that is for graphs with variable-nodes and !-boxes. This is
related to compositionality of the underlying concrete graphs by the
following instantiation homomorphism:

$\llbracket G \cric H \rrbracket = \llbracket G \rrbracket \cric \llbracket H \rrbracket$
